### 2025-08-29 sections/1-introduction.tex
found a of space

Examples of such datasets include genomic databases, time-series signals, and neural network embeddings such as GPT~\cite{2020arXiv200514165B, OpenAI2023GPT4TR}, LLAMA-2~\cite{Touvron2023Llama2O}, and image embedding models~\cite{radford2021learning, dosovitskiy2020image}.
Among biological datasets,  SILVA 18S~\cite{10.1093/nar/gks1219} includes ribosomal RNA sequences from approximately 2.25 million genomes, reaching 50,000 letters when aligned.
Among time-series datasets, the RadioML dataset~\cite{oshea2018radioml} contains approximately 2.55 million samples of synthetically-generated radio frequency signals.

@@

Given some measure of similarity between data points, two standard similarity search paradigms are $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN).
The former retrieves the $k$ points closest to a query, while the latter retrieves all points within a similarity threshold $\rho$ of a query.
While prior work has used the term \textit{approximate} search to refer to $\rho$-NN, we use it more narrowly: an {approximate} search algorithm is any algorithm that yields imperfect recall, whereas an \textit{exact} algorithm guarantees perfect recall.

$k$-NN search is one of the most widely-used methods in classification and recommendation~\cite{fix1952discriminatory, cover1967nearest}, but na\"{i}ve implementations are prohibitively slow because they scale linearly with dataset size.
Existing fast algorithms are often approximate~\cite{gao2023high}, which may suffice for some applications, but the need for efficient and \textit{exact} search remains~\cite{ukey2023survey}.
In a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but be sensitive to local perturbations for smaller values of $k$.

@@

Several algorithms have been developed to scale $k$-nearest neighbor search to large datasets, including Hierarchical Navigable Small World networks (HNSW)~\cite{malkov2016hnsw}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, random projection and tree building (ANNOY)~\cite{annoy}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. However, many of these algorithms do not support \emph{exact} search (as defined in Section \ref{sec:introduction}).

Hierarchical Navigable Small World networks~\cite{malkov2016hnsw} rely on navigable small world (NSW) networks and skip lists. InVerted File indexing (IVF)~\cite{faissivf, sacks1987multikey, kent1990signature} partitions data into high-dimensional Voronoi cells, and searches exhaustively only within cells near the query. ANNOY~\cite{annoy} uses random projection and tree building.
At each intermediate node of the tree, a random hyperplane, defined by two sampled points, splits the space.  Increasing the number of such trees improves recall but increases search time.

@@

Entropy-scaling algorithms exploit the inherent structure of large datasets, achieving complexity that scales with topological properties—such as metric entropy and fractal dimension—rather than cardinality. This makes them especially suitable for manifold-structured data.

In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search)~\cite{ishaq2019clustered}, which extended the flat clustering approach used in~\cite{yu2015entropy} to a hierarchical one.
CLAM (Clustering, Learning and Approximation with Manifolds), originally developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, refined the clustering algorithm from CHESS.

This paper introduces CAKES, a set of three entropy-scaling algorithms for $k$-NN search, implemented in Rust. We also provide a theoretical analysis of their time complexity in Sections~\ref{sec:methods:knn-search:repeated-rnn-complexity} and~\ref{sec:methods:knn-search:complexity-of-sieve-methods}.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a uniform distribution.
Given that CAKES's algorithms are intended for datasets with a manifold structure, complexity analyses assuming a uniform distribution would be uninformative.
As a result, our analysis assumes datasets with manifold structure—characterized by low fractal dimension and metric entropy—reflecting the conditions for which CAKES is designed.

>>>>

Examples of such datasets include genomic databases, time-series signals, and embeddings from neural networks such as GPT~\cite{2020arXiv200514165B, OpenAI2023GPT4TR}, LLAMA-2~\cite{Touvron2023Llama2O}, and image analysis models~\cite{radford2021learning, dosovitskiy2020image}.
Among biological datasets, SILVA 18S~\cite{10.1093/nar/gks1219} includes ribosomal RNA sequences from approximately 2.25 million genomes.
Among time-series datasets, the RadioML dataset~\cite{oshea2018radioml} contains approximately 2.55 million samples of synthetically-generated radio frequency signals.

@@

Given some measure of similarity between data points, two common similarity search paradigms are $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN).
The former retrieves the $k$ points closest to a query, while the latter retrieves all points within a similarity threshold $\rho$ of a query.
While prior work has used the term \textit{approximate} search to refer to $\rho$-NN, we use it more narrowly: an {approximate} search algorithm is any algorithm that yields imperfect recall, whereas an \textit{exact} algorithm guarantees perfect recall.

$k$-NN search is one of the most widely-used methods in classification and recommendation~\cite{fix1952discriminatory, cover1967nearest}, but na\"{i}ve implementations are prohibitively slow because they scale linearly with dataset size.
Existing fast algorithms are often approximate~\cite{gao2023high}, which may suffice for some applications, but the need for efficient and \textit{exact} search remains~\cite{ukey2023survey}.

@@

Several algorithms have been developed to tackle $k$-nearest neighbor search on large datasets, including Hierarchical Navigable Small World networks (HNSW)~\cite{malkov2016hnsw}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, random projection and tree building (ANNOY)~\cite{annoy}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}.
However, many of these algorithms do not support \emph{exact} search (as defined in Section \ref{sec:introduction}).

Hierarchical Navigable Small World networks~\cite{malkov2016hnsw} rely on navigable small world (NSW) networks and skip lists.
InVerted File indexing (IVF)~\cite{faissivf, sacks1987multikey, kent1990signature} partitions data into high-dimensional Voronoi cells, and searches only within cells near the query.
ANNOY~\cite{annoy} uses random projection and tree building.
At each intermediate node of the tree, a random hyperplane, defined by two sampled points, splits the space.
Increasing the number of such trees improves recall but increases search time.

@@

Entropy-scaling algorithms exploit the inherent structure of large datasets, achieving complexity that scales with topological properties, such as metric entropy and fractal dimension, rather than cardinality.
This makes them especially suitable for manifold-constrained data.

In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search)~\cite{ishaq2019clustered}, which extended the flat clustering from~\cite{yu2015entropy} to a hierarchical approach.
CLAM (Clustering, Learning and Approximation with Manifolds), originally developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, refines the clustering algorithm from CHESS.

This paper introduces CAKES, a set of three entropy-scaling algorithms for $k$-NN search, implemented in Rust.
We provide a theoretical analysis of their time complexity in Sections~\ref{sec:methods:knn-search:repeated-rnn-complexity} and~\ref{sec:methods:knn-search:complexity-of-sieve-methods}.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a uniform distribution.
Given that CAKES's algorithms are intended for datasets with a manifold structure, complexity analyses assuming a uniform distribution would be uninformative.
As a result, our analysis assumes datasets with manifold structure—characterized by low fractal dimension and metric entropy—reflecting the conditions for which CAKES is designed.
